The impact of earthquake cycle variability on neotectonic and paleoseismic slip rate estimates

### 2018-10-14 manuscript/reviews/response_letter.md

I performed this experiment; the figure with the slip rate results is shown
below for comparison with Figure 5 in the paper, and are included in a new
document in the supplemental material. The most relevant figure is also shown
here:

![Figure \label{fig}](../figures/nsv_slip_rate_envelopes.pdf)

The differences between the results of this experiment (fixed per-event
displacement of 1 m) and of the numerical experiment performed in the manuscript
are basically that these results are less smooth, but the total variance at any
point in time (*x-axis*) is less. This is simply due to removing the
stochasticity from one of the two variables in the system. The relative
spread in the data and convergence rates are unchanged. This is to be expected
as the variability in the per-event displacement is the same for all recurrence
distributions, so even though it is a random variable in the simulations, it is
not an experimental variable.

What I find the most interesting about this experiment is that the fluctuations
in the estimated slip rates show very clearly the mean earthquake cycles once
the noise from the per-event displacements has been removed. It is clear that
these are kind of damped or averaged out after ~7 earthquake cycles.
Nonetheless, though this is a cool pattern to see, I don't think it adds
enough insight to be worth including in the manuscript. I have added the figure
and table that show the slip rate variation through time to a new supplemental
document, accompanied by a brief discussion.

@@

> [O3] I would also like to a see a more quantitative comparison of COV and a 
> convergence on the mean to back up the assertion that COV of the distribution 
> is more important than the distribution itself.

This comment is addressed in the response to [O18] below.

> [O4] The paper would be improved by a more quantitative, empirical basis and 
> discussion of physical processes that may drive such recurrence behavior.

This paper is written to be a quick and clean demonstration of the effects of 
recurrence (and slip) variability on slip rate estimations. I want to keep the
work as simple and as digestible as possible, for a few reasons:

1. The intended audience for the paper is not only crustal deformation
   researchers, but others in the seismic hazard community as well--this
   includes engineers, geotechnical workers, analysts in the insurance industry,
   etc. In my experience as a member of this community, many others are
   only interested in these sorts of phenomena to the degree that they are
   consequential and actionable; their intellectual interests are often oriented
   towards their fields of expertise (structural engineering, ground motions,
   human and economic exposure, etc.). I want this paper to be a straightforward
   reference for how to evaluate slip rate data in light of aleatory variability
   that is not tied down in jargon or linked to specific geological or
   geophysical models or ideas that may not stand the test of time. Because of
   how variable and poorly-understood earthquake recurrence and fault
   interaction phenomena are, an in-depth discussion without resolution may well
   be off-putting to much of the audience that I would like to read this paper.

2. I don't believe that the state of the science is such that several paragraphs
   to a page of discussion can summarize our (minimal) understanding of these
   issues with any degree of satisfaction. Understanding earthquake recurrence
   and fault interaction is a fascinating research topic, and it deserves either
   a very full treatment (a review paper, lots of data analysis and modeling) or
   simply a few references before the paper moves on. We are not at the stage
   where we have a lot of answers, so it is very easy to get bogged down.

@@

> [O5] There is a literature of ideas to draw upon, such as post-seismic fault 
> reloading (Kenner and Simons, 2005), earthquake super cycles (Sieh et al., 
> 2008; Weldon et al., 2004), isolated versus fault-network behavior (Berryman et 
> al., 2012). Some of these ideas are discussed briefly but need more 
> explanation.

I disagree. It is clear that this stochasticity exists in nature, but I don't
think that we understand why. The modeling by Brad Hager's and Mark Simons'
research groups

@@

> [O6a] Likewise one could examine actual earthquake slip distributions (not 
> landform offsets of historic events, which convolve landscape processes with 
> tectonic slip) to develop an empirical basis for the slip function.

Such a distribution is given by Biasi and Weldon (2009), following work done by 
Hemphill-Haley and Weldon (1999). It is a bit different than the lognormal 
distribution used in that the probability of relatively low values (zero or 
near-zero) is higher than in a lognormal distribution. The sample COV of this 
distribution is 0.67, slightly lower than the lognormal slip distribution used
in the paper, with a COV of 0.75.

As an experiment, I have re-done the simulation sampling from this 
distribution; the data are given as 1313 discrete points from earthquakes 
worldwide, normalized to the mean slip per event. I have sampled randomly from 
this finite set, with replacement, instead of interpolating the set into a 
continuous distribution and sampling from that. The results are in the new
supplemental materials (Figures S1 and S2, Table S1, and some discussion).

The results are nearly indistinguishable.

@@

*No changes.*

> [O18] Page 5, line 15. The statement ‘appears to be related to COV’ is 
> disappointing. Given that this paper is entirely simulation, the author 
> should be able to make a quantitative comparison of slip-rate variance to 
> COV.

> [O19] Page 5, line 29. This problem has been studied (Weldon et al., 2004; 
> Sieh et al., 2008). The Sumatran subduction zone work is particularly 
> relevant and completely overlooked here.

The work by Weldon et al. (2004) is referenced on Page 6, Line 3. Neither of
these papers deal with the topic of autocorrelated recurrence intervals in any
quantitative or otherwise explicit manner, and I don't read anything that I can
interpret as a qualitative discussion either. One can read both papers and not
get a sense of whether a short recurrence interval implies the next recurrence
interval will be short or long, much less any quantification. Both papers deal
with the concept of 'earthquake supercycles' or groups of earthquakes that are
relatively tightly-spaced and separated from other groups by long recurrence
intervals. This may share a conceptual link, but from a technical perspective
this likely has more to do with periodicity than autocorrelation, and these are
mathematically separate such that a periodic sequence may have positive or
negative autocorrelation. Goh and Barabasi (2008, Europhysics Letters) is a
useful discussion on the topic of autocorrelation vs. periodicity in regards to
quantifying clustering behavior (though one may safely ignore the references to
seismicity in that paper).

[Note: I've analyzed Weldon's data (from K. Scharer's refined OxCal earthquake 
dates) and found that earthquake recurrence interval duration at the Wrightwood 
and Pallett Creek sites has a negative autocorrelation, i.e. a short recurrence 
interval is likely to be followed by a long recurrence interval, and vice 
versa. This is quite unlike the autocorrelation in the Puget Lowlands of
Washington State, which is more like a network of generally similar faults
instead of an isolated or dominant plate boundary fault like the SAF or the
Alpine Fault as referenced in another comment on Kelvin Berryman's work, where a
long recurrence interval will more likely be followed by another long recurrence
interval. I don't know what this means, or whether it's all noise, but it's
intriguing to me and I put this bit in the paper in hopes of catching the
attention of others who may be looking for a project.]

@@

> [O20] Page 6, line 1. Zero friction at rupture arrest is very unrealistic, 
> and not a prerequisite for characteristic behavior.

It is not a necessary condition but it is a sufficient one, when coupled with 
fairly regular reloading and failure conditions. I personally think it's 
unlikely as well (see Styron and Hetland 2015 for example), but zero (or very
near zero) friction is part and parcel with complete stress drop in major
earthquakes, which is supported by many studies [REFS]. Furthermore, there are a
host of laboratory experiments which suggest that friction during slip decreases
to very low values [MORE REFS].

### 2018-10-14 manuscript/eq_cycle_slip_rates.tex
updated all mods

Fault slip rates are generally estimated by dividing measurements of the
offset of geologic `marker' features by the time over which that offset
accumulated. The uncertainty in the resulting slip estimate is typically
treated as \emph{epistemic}, and quantified through the propagation of
the measurement uncertainties on the offset and time quantities
\citep[e.g.]{bird_uncertainties_2007,zechar_incorporating_2009}.
However, for slip rate estimates on active faults made from offset
measurements near the fault trace (i.e., within a horizontal distance
that is a small fraction of the fault's locking depth), the episodic
nature of surface displacement due to the earthquake cycle will
necessarily affect the results: If the measurements are taken
immediately before an earthquake, the measured offset and resulting slip
rate estimate will be lower than average, while if the measurements are
taken immediately after an earthquake, the offset and rate will be
higher.

@@

However, the recurrence intervals between successive earthquakes on any
given fault segment have some natural variability (i.e. \emph{aleatoric
uncertainty}); similarly, displacement at a measured point is not
identical in each earthquake \citep[e.g.]{duross_holocene_2008}.
Therefore, the measured slip rate may deviate from the time-averaged
rate based on the amount of natural variability in the earthquake cycle,
particularly given successive events from the tails of the recurrence
interval or displacement distributions. Though the physical mechanisms
responsible and the statistical character of this natural variability
remain under debate, its effects on the estimated slip rates may still
be estimated given some common parameterizations.

@@

To isolate the effects of the earthquake cycle from other phenomena that
may affect slip rate estimates, this study does not attempt to model
erosion, nor does it consider any measurement uncertainty in the age or
offset of the faulted geologic markers; these measurements are assumed
to be perfect. Additionally, though natural variability in
per-earthquake displacement is included in the model, it is minor and
the same for all recurrence distributions. Furthermore, though the model
has one length dimension (fault offset), it is still best thought of as
a point (0-dimensional) model, as there is no spatial reference or
along-strike or down-dip variability, and hence the magnitude of each
earthquake is undefined, and no magnitude-frequency distribution exists.

@@

The choice of the lognormal distribution is for convenience, simplicity, and
flexibility: it is a common, well-known distribution and--should one be
interested--can be easily given different shape and scale values to modify the
\(COV\) or change the mean slip rate in the modeling code used in this paper.
However, it is not necessarily the most accurate representation of earthquake
slip variability. **BIASI AND WELDON (2006)** compiled field measurements of
surface ruptures from 13 earthquakes, and the resulting distribution has some
significant differences with the lognormal distribution used here, though the
\(COV\) of 0.67 is quite close to the value used here. To test the
sensitivity of the results given in this paper to the choice of slip
distribution, the numerical simulations in presented in this work were run with
the only change being the use of the empirical slip distribution from **BIASI
AND WELDON (2006)**, and the results are given in the supplemental material.
Though there is more discussion there, the results are essentially identical to
those presented below.

>>>>

Fault slip rates are generally estimated by dividing measurements of the
offset of geologic marker features by the time over which that offset
accumulated. The uncertainty in the resulting slip estimate is typically
treated as \emph{epistemic}, and quantified through the propagation of
the measurement uncertainties on the offset and time quantities
\citep[e.g.]{bird_uncertainties_2007,zechar_incorporating_2009}.
However, for slip rate estimates on active faults made from offset
measurements near the fault trace (i.e., within a horizontal distance
that is a small fraction of the fault's locking depth), the episodic
nature of surface displacement due to the earthquake cycle will
necessarily affect the results: If the measurements are taken
immediately before an earthquake, the measured offset and resulting slip
rate estimate will be lower than average, while if the measurements are
taken immediately after an earthquake, the offset and rate will be
higher.

@@

However, the recurrence intervals between successive earthquakes on any
given fault segment have some natural variability (i.e. \emph{aleatoric
uncertainty}); similarly, displacement at a measured point is not
identical in each earthquake \citep[e.g.]{duross_holocene_2008}.
Therefore, the measured slip rate may deviate from the time-averaged
rate based on the amount of natural variability in the earthquake cycle,
particularly given successive events from the tails of the recurrence
interval or displacement distributions.

The physical mechanisms responsible for the aleatoric variability in earthquake
recurrence intervals and displacements are still unclear, and the subject of
active investigation. Most earthquakes serve to release differential stresses
caused by relative motions of tectonic plates or smaller crustal blocks;
relative plate velocities measured over tens of thousands to millions of years
from geologic reconstructions are similar enough to those measured over a few
years through GPS geodesy that sudden, transient accelerations and decelerations
are unlikely (e.g., **DEMETS AND DIXON 1999**). As a consequence, some plate
boundary faults are among the most regularly-rupturing faults known,
particularly sections that are isolated from nearby faults
\citep[e.g.]{berryman_major_2012} and therefore not affected by stress
perturbations resulting from earthquakes on other faults. These stress changes
may be 'static' coseismic instantaneous stresses in the elastic upper crust resulting from
earthquake displacement, or analogous stress changes in the viscoelastic lower
crust or upper mantle from the time-dependent relaxation of static stress
perturbations, or 'dynamic' transient stress changes that accompany the passage
of seismic waves from nearby or distant earthquakes. Regardless of the nature

Intraplate faults

It is important to keep in mind that each earthquake in a sequence places
points of rock on both sides of the fault that were not previously in contact.

Though the physical mechanisms
responsible and the statistical character of this natural variability
remain under debate, its effects on the estimated slip rates may still
be estimated given some common parameterizations.

@@

To isolate the effects of the earthquake cycle from other phenomena that
may affect slip rate estimates, this study does not attempt to model
erosion, nor does it consider any measurement uncertainty in the age or
offset of the faulted geologic markers; these quantities are assumed
to be perfectly known. Additionally, though natural variability in
per-earthquake displacement is included in the model, it is minor and
the same for all recurrence distributions; though it is a random variable in the
simulations, it is not an experimental variable. Furthermore, though the model
has one length dimension (fault offset), it is still best thought of as
a point (0-dimensional) model, as there is no spatial reference or
along-strike or down-dip variability, and hence the magnitude of each
earthquake is undefined, and no magnitude-frequency distribution exists.

@@

The choice of the lognormal distribution is for convenience, simplicity, and
flexibility: it is a common, well-known distribution and--should one be
interested--can be easily given different shape and scale values to modify the
\(COV\) or change the mean slip rate in the modeling code used in this paper.

However, it is not necessarily the most accurate representation of earthquake
slip variability. **BIASI AND WELDON (2006)** compiled field measurements of
surface ruptures from 13 earthquakes. The resulting distribution (Figure S1) has
some significant differences with the lognormal distribution used here, though
the \(COV\) of 0.67 is quite close to the value used here. To test the
sensitivity of the results given in this paper to the choice of slip
distribution, the numerical simulations in presented in this work were run with
the only change being the use of the empirical slip distribution from **BIASI
AND WELDON (2006)**, and the results are given in the supplemental material
(Figure S2, Table S1). Though there is more discussion in the supplements, the
results are essentially identical to those presented below. As an additional
experiment, the numerical simulations have been run using an invariant per-event
displacement of 1 m. Though this is not a realistic scenario, it allows for a
deconvolution of the effects of earthquake time stochasticity and earthquake
displacement stochasticity. The results are shown in Figure S3 and Table S2, and
discussed in the supplemental materials; there are noticeable differences in the
results, but they are small and do not call into question the results and
conclusions presented below.
